---
title: "Homework 1 - readable and efficient R code"
author: "Yating Zou"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

```{r}
# Load packages needed
library(microbenchmark)
```

# Question 1 - "function-alize" this code

Read over the code below and perform the following:

* Wrap it into a function `foobar0` which has arguments `x` and `z`
  and which returns the vector `x` at the end of the following code.
* Rewrite this into a function `foobar` which is easier to read, by
  reducing repetitive code. E.g. `foobar` might call a function to
  check the input, and another function to perform the three lines of
  computation. 
* Check that the two versions produce the same output using the
  function `all.equal`.

```{r}
# Function foobar0 =================================
foobar0 <- function(x, z){
  
  if (sum(x >= .001) < 1) {
    stop("step 1 requires 1 observation(s) with value >= .001")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- sin(r) + .01
  
  if (sum(x >= .002) < 2) {
    stop("step 2 requires 2 observation(s) with value >= .002")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 2 * sin(r) + .02
  
  if (sum(x >= .003) < 3) {
    stop("step 3 requires 3 observation(s) with value >= .003")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 3 * sin(r) + .03
  
  if (sum(x >= .004) < 4) {
    stop("step 4 requires 4 observation(s) with value >= .004")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 4 * sin(r) + .04

return(x)
}


# Function foobar =================================

## helper function for checking the input
check_input <- function(x, c){
  if (sum(x >= c/10^3) < c){
    stop(cat("step", c, "requires",c, "observation(s) with value >=", c/10^3))
  }
}

## helper function for computing x at step c
compute <- function(x, z, c) {
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- c * sin(r) + c / 10 ^ 2
  return(x)
}

# main function foobar
foobar <- function(x, z) {
  for (c in 1:4) {
    check_input(x, c)
    x <- compute(x, z, c)
  }
  return(x)
}
```

Check if `foobar0()` and `foobar()` are equal using `base::all.equal()`.
```{r}
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)

all.equal(foobar0(x, z), foobar(x, z))
```


# Question 2 - vectorize this code and benchmark

* Take the following function `f0` and rewrite it as a function `f`,
  which is faster and easier to read, by removing the loop of `i` from
  1 to `m`.
* Benchmark `f` and `f0` using `microbenchmark`. How much faster is `f`?

```{r}
# Function using for-loop ==============================
f0 <- function(x, b, a) {
  out <- numeric(0)
  for (i in seq_len(m)) {
    bb <- b[i,]
    aa <- a[i,]
    out <- c(out, sum(x %*% bb) + sum(aa))
  }
  out
}


# Function using vectors ================================
f <- function(x, b, a) {
  out <- colSums(x %*% t(b)) + rowSums(a)
  out
}
```

Benchmark `f0()` and `f()` using `microbenchmark::microbenchmark()`.

```{r}
n <- 30
p <- 50
p2 <- 25
m <- 1000
set.seed(1)
x <- matrix(rnorm(n * p), nrow = n, ncol = p)
b <- matrix(rnorm(m * p), nrow = m, ncol = p)
a <- matrix(rnorm(m * p2), nrow = m, ncol = p2)

microbenchmark(f0(x, b, a), f(x, b, a))
```

The run-time of `f0()` is `r round(5.9/1.556,2)` times that of `f()` on average.


# Question 3 - build a faster t-test

* Rewrite the following function `getT0` which computes `m`
  two-sample t-tests (equal variance) between two groups as a function
  `getT`, which is faster by using vectorized operations over the `m`
  sets of observations. (There are functions in R packages, such as
  `genefilter::rowttests` which will quickly perform this operation,
  but I want you to write your own function using simple R functions
  like `rowSums`, etc.)
* Benchmark `getT` and `getT0`. How much faster is `getT`?

```{r}
# Function using sapply ===============================
getT0 <- function(x, f) {
  ts <-
    sapply(seq_len(m), function(i)
      t.test(x[i, ] ~ f, var.equal = TRUE)$statistic)
  unname(ts)
}


# Function using vectorization =========================
getT <- function(x, f) {
  n1 <- sum(f == 1)
  n2 <- sum(f == 2)
  s1.squareds <- rowSums((x[f == 1,] - rowMeans(x[f == 1, ])) ^ 2)
  s2.squareds <- rowSums((x[f == 2,] - rowMeans(x[f == 2, ])) ^ 2)
  
  diffs <- rowMeans(x[f == 1, ]) - rowMeans(x[f == 2, ])
  s.squareds <-
    ((n1 - 1) * s1.squareds + (n2 - 1) * s2.squareds) / (n1 + n2 - 2)
  
  ts <- diffs / sqrt(2 * s.squareds / little.n)
  ts
}
```

Benchmark `getT0()` and `getT()` using `microbenchmark::microbenchmark()`.
```{r}
m <- 400
n <- 50
little.n <- n / 2
set.seed(1)
x <- matrix(rnorm(m * n), nrow = m, ncol = n)
f <- gl(2, little.n)

microbenchmark(getT0(x, f), getT(x, f))
```

The run-time of `getT0()` is `r round(172660/867, 2)` times that of `getT()` on average.

